HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations

### 2021-03-25 docs/main.tex
small changes in text

HAMdetector shows strong support for 15 replacement - allele pairs that have previously not been identified. All of them exceed the significance level of 
$5\times 10^{-3}$ used in~\citet{Karimzadeh2019}. This shows that HAMdetector is able to identify possible HAMs that would not be detected otherwise. The most striking example for this is K43R - A*02, which using Fisher's exact test has a p-value of 0.22. While this replacement has not been experimentally verified in a IFN-\(\gamma\) production assay yet, it lies within a known A*02 epitope and thus it is highly likely that K43R conveys some form of immune escape.

\begin{table}
  \caption{List of HAMs predicted Fisher's exact test (FET). In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation. Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector. \emph{Bona fide} predictions are separated by a horizontal line. These are replacement - allele pairs we expect to to be verified in future experimental tests. The replacement annotated with a (+) lies within a known A*02 epitope.}
  % \vspace{0.5cm}
  \begin{tabular}{c|c|c|c|c}
  replacement & allele  & p-value (FET) & post.~prob. & confirmed \\
  \hline
  S170N       & B*15 & $3\cdot10^{-8}$      & 0.99                                & +                        \\
  D101E       & B*37 & 0.0002                        & 0.96                                & +                        \\
  R105K       & B*27 & 0.0011                        & 0.93                                & +                        \\
  R139K       & B*41 & 0.0034                        & 0.92                                & +                        \\
  E47D        & B*18 & 0.0027                        & 0.90                                & +                        \\
  D33E        & B*13 & 0.0001                        & 0.86                                & -                        \\
  T134A       & A*68 & 0.0045                        & 0.82                                & -                        \\
  K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
  D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
  K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
  A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
  P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
  Q100L       & B*13 & 0.0018                        & 0.60                                & -                        \\
  D96E        & B*13 & 0.0035                        & 0.51                                & -  \\\hline
  E46D & A*02 & 0.0054 & 0.97 & ? \\
  V81I & A*68 & 0.0073 & 0.97 & ? \\
  K113N & B*08 & 0.0063 & 0.96 & ? \\
  A71T & B*41 & 0.0065 & 0.96 & ? \\
  L188I & A*68 & 0.0632 & 0.94 & ? \\
  T95S & A*01 & 0.0285 & 0.93 & ? \\
  D33E & A*03 & 0.0226 & 0.93 & ? \\
  P49L & B*13 & 0.0035 & 0.92 & ? \\
  A74S & A*68 & 0.0123 & 0.91 & ? \\
  E29D & B*44 & 0.0559 & 0.91 & ? \\
  D46E & B*57 & 0.0190 & 0.91 & ? \\
  R88K & A*68 & 0.0123 & 0.91 & ? \\
  T149P & B*52 & 0.0281 & 0.91 & ? \\
  K43R & A*02 & 0.2158 & 0.90 & (+) \\
  N22S & B*08 & 0.0405 & 0.90 & ? \\
  \end{tabular}
  \label{tab:false-positives}
 \end{table}

@@

 Table~\ref{tab:false-positives} shows that for 3 of the false positives identified with Fisher's exact test, HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03).
 One possible explanation for this observation is HLA-Linkage disequilibrium and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)). A similar observation can be done for K43R and D33E, albeit the correlation between the respective alleles is much weaker.

\begin{figure}
  \includegraphics[width=1.1\linewidth]{plots/P49L_combined.pdf}
  \caption{Marginal posteriors for the regression coefficients of A*30 and B*13 for replacement P49L. The association between P49L and A*30 identified by Fisher's exact test turned out to be a false positive.}
  \label{fig:linkage}
\end{figure}

Figure~\ref{fig:linkage} shows regression coefficients of the HLA alleles A*30 and B*13 for replacement P49L. When using a simple logistic regression model without any additions (model~1), both A*30 and B*13 show medium evidence of being associated with replacement P49L. However, when also taking phylogeny into account and using a sparsity-promoting prior (model~3), both regression coefficients get shrunk to 0 because the association is not strongly enough identified by the data alone.
Using epitope prediction as an additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as associated with P49L. The false-positive association between P49L and A*30 remains being shrunk towards 0.  
\section{Discussion}

>>>>

HAMdetector strongly supports 15 replacement - allele pairs that have previously not been identified (question marks in last column of Table~\ref{tab:false-positives}). All of them have association probabilities of 0.90 or higher, while their p-values from Fisher's exact test exceed the significance level of $5\times 10^{-3}$ used in~\citet{Karimzadeh2019}. Given the superior performance of HAMdetector on the experimentally tested HAMs, these 15 bona fide predictions suggest that most true HAMs may still to be discovered. A striking example is K43R - A*02 with a p-value of 0.22 in Fisher's exact test but a HAM-probability of 0.90 and location inside an A*02 restricted epitope.

\begin{table}[h!]
  \caption{List of HAMs predicted by Fisher's exact test (FET). In the last column ``+'' and ``-'' mark experimentally confirmed or rejected HAMs, respectively; ``?'' below the horizontal line indicate untested bona fide predictions. ``post.prob.'' are posterior probabilities for positive associations computed with HAMdetector.}
  \vspace{0.5cm}
  \begin{tabular}{c|c|c|c|c}
  replacement & allele  & p-value (FET) & post.~prob. & confirmed \\
  \hline
  S170N       & B*15 & $3\cdot10^{-8}$      & 0.99                                & +                        \\
  D101E       & B*37 & 0.0002                        & 0.96                                & +                        \\
  R105K       & B*27 & 0.0011                        & 0.93                                & +                        \\
  R139K       & B*41 & 0.0034                        & 0.92                                & +                        \\
  E47D        & B*18 & 0.0027                        & 0.90                                & +                        \\
  D33E        & B*13 & 0.0001                        & 0.86                                & -                        \\
  T134A       & A*68 & 0.0045                        & 0.82                                & -                        \\
  K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
  D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
  K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
  A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
  P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
  Q100L       & B*13 & 0.0018                        & 0.60                                & -                        \\
  D96E        & B*13 & 0.0035                        & 0.51                                & -  \\\hline
  E46D & A*02 & 0.0054 & 0.97 & ? \\
  V81I & A*68 & 0.0073 & 0.97 & ? \\
  K113N & B*08 & 0.0063 & 0.96 & ? \\
  A71T & B*41 & 0.0065 & 0.96 & ? \\
  L188I & A*68 & 0.0632 & 0.94 & ? \\
  T95S & A*01 & 0.0285 & 0.93 & ? \\
  D33E & A*03 & 0.0226 & 0.93 & ? \\
  P49L & B*13 & 0.0035 & 0.92 & ? \\
  A74S & A*68 & 0.0123 & 0.91 & ? \\
  E29D & B*44 & 0.0559 & 0.91 & ? \\
  D46E & B*57 & 0.0190 & 0.91 & ? \\
  R88K & A*68 & 0.0123 & 0.91 & ? \\
  T149P & B*52 & 0.0281 & 0.91 & ? \\
  K43R & A*02 & 0.2158 & 0.90 & ? \\
  N22S & B*08 & 0.0405 & 0.90 & ? \\
  \end{tabular}
  \label{tab:false-positives}
 \end{table}

@@

 For three of the false positives proposed by Fisher's exact test (Table~\ref{tab:false-positives}), HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03). One possible explanation for this observation is HLA linkage disequilibrium \hl{DaHo: and accordingly correlation between HLA alleles in the dataset due to random sampling variation. (Satz etwas opak.)} Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)). A similar observation can be made for K43R and D33E, although the correlation between the respective alleles is much weaker.

\begin{figure}
  \includegraphics[width=1.1\linewidth]{plots/P49L_combined.pdf}
  \caption{Marginal posteriors for the regression coefficients of A*30 (left column) and B*13 (right column) for replacement P49L with different model stages (rows).}
  \label{fig:linkage}
\end{figure}

Figure~\ref{fig:linkage} shows regression coefficients of the HLA alleles A*30 and B*13 for replacement P49L. With the simplest logistic regression model (model~1), both A*30 and B*13 have medium evidence of being associated with replacement P49L. However, with phylogeny and sparsity-promoting prior (model~3) both regression coefficients shrink close to 0 -- the associations are not convincingly supported by the data. Using epitope prediction as additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as likely associated with P49L. The association between P49L and A*30 (predicted by Fisher's exact test) remains shrunk towards 0.  
\section{Discussion}
